---
title: "Data Wrangling"
author: "Melinda K. Higgins, PhD."
date: "September 11, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

## Data Wrangling

For today's lesson, you will want to download/make a copy of the Github repository [https://github.com/melindahiggins2000/N736Fall2017_lesson06](https://github.com/melindahiggins2000/N736Fall2017_lesson06). This repo contains 3 datasets we will be working with in the exercises below: 

* `dataA.csv` - has 3 variables (var1, var2, var3) and 6 subjects (1,2,3,5,8,9)
* `dataB.csv` - has 3 variables: 2 same and 1 different than `dataA` (var1, var2, var4) and 6 more subjects (11,12,14,15,17,18)
* `dataC.csv` - has 3 more variables (var5, var6, var7) and 6 subjects (1,2,3,4,7,10) some in common with `dataA` and some different

Lesson 05 covered:

* merging; 
* joining; 
* stacking; and 
* concatenating 2 datasets together.

For Lesson 06 we'll cover:

* adding labels and other notations
* recoding

In future lessons, we'll also discuss:

* manipulating data types
* switching from numeric to character and back
* handling dates and times
* handling `factors` in `R` specifically - but in SAS (and in SPSS), numeric variables assumed to be continuous are modeled differently than numeric variables treated as "factors" (i.e. ordinal or categorical data represented numerically).

## Adding Labels and Other Notations

It can sometimes be helpful to add a label to your dataset or to your variables.

In `R` you can add a comment on your dataset which can then be pulled up using the `attributes()` function, which could be used later in a title or other text function.

**R CODE**

```{r, echo=TRUE}
library(readr)
library(knitr)
dataA <- read_csv("dataA.csv")
dataB <- read_csv("dataB.csv")

comment(dataA) <- "This is a comment on dataset dataA"
comment(dataB) <- "This is a comment on dataset dataB"

attributes(dataA)
attributes(dataB)
```

I could use the result of the `attributes()` function to print out info on `dataA`.

```{r, echo=TRUE}
print(paste("Here are the details on dataset dataA:",
            attributes(dataA)$comment))
```

**SAS Code**

```{r}
library(SASmarkdown)

saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -linesize 75"

knitr::opts_chunk$set(engine="sashtml", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)

sas_collectcode()
```

Load `dataA` and `dataB`.

```{r, echo=TRUE, collectcode=TRUE}
* ======================================;
* load dataA;
* ======================================;
proc import datafile='C:\MyGithub\N736Fall2017_lesson05\dataA.csv'
  out=dataA dbms=csv;
run;

* ======================================;
* load dataB;  
* ======================================;
proc import datafile='C:\MyGithub\N736Fall2017_lesson05\dataB.csv'
  out=dataB dbms=csv;
run;
```

In `SAS` it is possible to add a comment about the dataset, but this only shows up as a note in output like `PROC CONTENTS`.

```{r, echo=TRUE, collectcode=TRUE}
data dataAcomment (label="This is a comment on dataA");
  set dataA;
run;

proc contents data=dataAcomment; run;
```

### adding labels and levels for `factors` in `R`

**R Code**

```{r, engine='R'}
# reset the engine globally (i.e. so it will work outside of
# this chunk so you don't have to keep typing engine=`R`)
knitr::opts_chunk$set(engine="R",  engine.path=NULL, 
        engine.opts=NULL, comment=NA)
# knitr::opts_chunk$get()$engine
``` 

For `var` which has 5-point LiKert scale data coded from 1 to 5, let's assume it should have the following labeling and interpretation:

```
1 = "strongly disagree"
2 = "disagree"
3 = "neutral"
4 = "agree"
5 = "strongly agree"
```

Also notice the changes in the "class"/"variable type" when we make `var1` a factor.

```{r, echo=TRUE}
kable(dataA)
class(dataA$var1)
typeof(dataA$var1)
is.numeric(dataA$var1)

dataA$var1 <- factor(dataA$var1,
                     levels=c(1,2,3,4,5),
                     labels=c("strongly disagree",
                              "disagree", "neutral",
                              "agree","strongly agree")) 

kable(dataA)
class(dataA$var1)
typeof(dataA$var1)
is.numeric(dataA$var1)
```

We can also apply better labeling for gender `var2` which is currently a character variable coed as "m" or "f".

```{r, echo=TRUE}
kable(dataA)
class(dataA$var2)
typeof(dataA$var2)
is.numeric(dataA$var2)

dataA$var2 <- factor(dataA$var2,
                     levels=c("m","f"),
                     labels=c("male","female")) 

kable(dataA)
class(dataA$var2)
typeof(dataA$var2)
is.numeric(dataA$var2)
```

### using `SAS` formats

**SAS Code**

```{r}
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -linesize 75"

knitr::opts_chunk$set(engine="sashtml", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)
```

Create some formats for likert scale variables and for the gender character variable.

```{r, echo=TRUE, collectcode=TRUE}
proc format;
  value likert  
    1='strongly disagree'
    2='disagree'
  	3='neutral'
	  4='agree'
	  5='strongly agree';
  value $gender  
    'm'='male'
    'f'='female';
run;
    
```

See what it looks like without the format applied.

```{r, echo=TRUE, collectcode=TRUE}
proc print data=dataA; run;
```

Apply the format to a new dataset `dataA2` for the `var1` variable.

```{r, echo=TRUE, collectcode=TRUE}
data dataA2;
  set dataA;
  format var1 likert.;
run;

proc print data=dataA2; run;
```

You don't have to use the format in a `DATA` step, you can simply use it when you need it, such as within a proc print statement to get the formatting you want in the output. Let's use the `format` statement here when we print the data table for `dataB`.

```{r, echo=TRUE, collectcode=TRUE}
proc print data=dataB;
  format var2 $gender.;
  run;
```

We can also use the `format` statement here to print the likert scale labels in a frequency table. First, without formatting and again with the formatting - compare the output.

```{r, echo=TRUE, collectcode=TRUE}
* run a frequency table without formatting;

proc freq data=dataB;
  tables var2;
  run;

* add the formatting for the table;

proc freq data=dataB;
  format var2 $gender.;
  tables var2;
  run;
```

### tables with labels

Frequency table and plots without `format` applied for `dataA` and again with `format` applied for `dataA2`.

**SAS Code**

```{r, echo=TRUE, collectcode=TRUE}
proc freq data=dataA;
 tables var1;
 run;

proc freq data=dataA2;
 tables var1;
 run;
```

### Other labeling in `SAS` - titles

Here is an example of adding a title to the ODS output.

```{r, echo=TRUE, collectcode=TRUE}
proc freq data=dataB;
  title "Frequency Table of Gender";
  format var2 $gender.;
  tables var2;
  run;
```

Notice the title remains in the next output also because the `title` command is a GLOBAL setting that stays in effect until you change it.

**NOTE** The changes will show up in the `SAS` output window but are not shown here in this `markdown` based output.

```{r, echo=TRUE, collectcode=TRUE}
proc print data=dataB; run;
```

Set the title back to empty. In general, it is a good idea to ALWAYS have some kind of title with your output to keep track.

```{r, echo=TRUE, collectcode=TRUE}
title;

proc print data=dataB; run;
```

You can also use footnotes. You can have up to 3 titles and 3 footnotes.

see more at [http://support.sas.com/documentation/cdl/en/lestmtsref/69738/HTML/default/viewer.htm#p10gcmrmf83iaxn1ilrx4pra969n.htm](http://support.sas.com/documentation/cdl/en/lestmtsref/69738/HTML/default/viewer.htm#p10gcmrmf83iaxn1ilrx4pra969n.htm).

```{r, echo=TRUE, collectcode=TRUE}
title "Univariate Report for var 1 in dataA2";
title2 "Report by Melinda Higgins";
title3 "dated 09/15/2017";
footnote "this is footnote 1";
footnote2 "this is footnote 2";
footnote3 "this is footnote 3";

proc univariate data=dataA2 plots;
  var var1;
  histogram var1/normal;
  run;

* set all back to empty;
title;
title2;
title3;
footnote;
footnote2;
footnote3;
```

```{r, engine='R'}
# reset the engine globally (i.e. so it will work outside of
# this chunk so you don't have to keep typing engine=`R`)
knitr::opts_chunk$set(engine="R",  engine.path=NULL, 
        engine.opts=NULL, comment=NA)
knitr::opts_chunk$get()$engine
```

## Recoding

Suppose we want to recode `var1` into a new variable identifying subjects with `var1` scores <=3 and those >3 (splitting the data at 3). We'll create the new variable and then add the formatting we want using the `factor()` function in `R`.

Since we made `var1` a "factor" type variable above, we have to use the `as.numeric()` function to be able to manipulate the numeric values underlying this variable as a factor.

We also use the `as.numeric()` function again since the statement `x > 3` is a logic statement and returns `FALSE` and `TRUE` and what we want is `0` and `1`. If you do want a logic variable you can leave off this 2nd (outer) `as.numeric()` function.

**R Code**

```{r, echo=TRUE}
x <- c(1,2,3,4,5)
x > 3
typeof(x > 3)

dataA$var1r <- as.numeric(
  as.numeric(dataA$var1) > 3
  )

typeof(dataA$var1r)

dataA$var1r <- factor(dataA$var1r,
                      levels=c(0,1),
                      labels=c("NO var1 <= 3",
                               "YES var1 > 3")) 

kable(dataA)
```

**SAS Code**

```{r}
saspath <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -linesize 75"

knitr::opts_chunk$set(engine="sashtml", engine.path=saspath, 
        engine.opts=sasopts, comment=NA)
```

From lesson 5 - merge `dataA` and `dataB` together.

```{r, echo=TRUE, collectcode=TRUE}
data dataAB;
  set dataA dataB;
run;

proc print data=dataAB noobs; run;
```

Let's recode `var1` into a new variable identifying subjects with `var1` scores <=3 and those >3 (splitting the data at 3). First we'll define the formatting we want for our new variable and after we recode it in the `DATA` step we'll apply the formatting.

```{r, echo=TRUE, collectcode=TRUE}
proc format;
  value splitYN
    0 = 'NO var1 <= 3'
    1 = 'YES var1 > 3';
run;
```

Create new dataset with recoded variables and apply new format labels.

```{r, echo=TRUE, collectcode=TRUE}
data dataAB_new;
  set dataAB;
  if var1 > 3 then var1split3 = 1;
  else             var1split3 = 0;
  format var1split3 splitYN.;
run;
```

Create table and plot of new variable split <=3 and >3.

```{r, echo=TRUE, collectcode=TRUE}
proc freq data=dataAB_new;
 tables var1split3/plots=freqplot;
 run;
```

```{r, engine='R'}
# reset the engine globally (i.e. so it will work outside of
# this chunk so you don't have to keep typing engine=`R`)
knitr::opts_chunk$set(engine="R",  engine.path=NULL, 
        engine.opts=NULL, comment=NA)
# knitr::opts_chunk$get()$engine
```
